A molecular taxonomy of tumors independent of tissue-of-origin

### 2020-08-24 text/manuscript.Rmd
add supplementary methods, move clustering and survival methods to supp methods doc, rename figure2.Rmd-->figureSM1.Rmd

In order to classify tumors using this dataset, we used multiple correspondence
analysis (MCA) [@JSSv025i01]. First, we determined the number of dimensions
containing useful information by selecting the eigenvalue with the most
explanatory power, using the average of 100 permutations of the data as baseline
(**figure 2A**). We then chose the maximum eigenvalue for which the p-value
remained ≤ 0.05 (see cutoff in **figure 2B**). Then we performed a UMAP
analysis, both in order to summarize the MCA graphically, and as a preprocessing
step to boost the performance of density based clustering. The resulting map was
notable for its lobed structure, with several reproducible projections
regardless of random seed setting. A representative version of this 3D UMAP is
shown in **figure 2C**, rotated to enhance the visibility of the major features.
Following this spatial mapping we attempted to define groupings of similar
tumors within the spatial map using HDBSCAN, which performs hierarchical
clustering and provides metrics of cluster stability and probabilities of
cluster membership for each node. However, HDBSCAN is sensitive to several
parameters; key for our analysis are the minimum number of tumor samples in a
cluster that capture the maximum number of tumor samples, measured by
probability of membership of ≥ 5% in at least one class. Thus, we created a
score metric as the fraction of classified tumors with max probability < 5% in
one class and chose a cluster size of 92 to minimize the score function
(**figure 2D**). HDBSCAN with these settings resulted in ten distinct
high-density clusters which we then projected onto the UMAP (**figure 2E**).
This classified 6,038 out of 7,607 tumors but still resulted in a significant
fraction of unclassified tumor samples. Since we ultimately wish to be able to
classify any tumor using this scheme, we performed k Nearest Neighbors (kNN)
analysis, which computes a similarity metric to every tumor in the set and then
lets the _k_ most similar tumors "vote" as to the identity of the query tumor
sample based on their classification labels. We set _k_ to be the square root of
the number of tumor samples (87). Using this method, we assigned cluster
membership to the remaining tumors (**figure 2F**).

@@

#### Figure 2
```{r child='../code/figure2.Rmd'}
```

**Figure 2: Method for tumor classification using pathway aberration data** 
a) MCA analysis was performed to determine the eigenvalue with the most
explanatory power from the categorical data. Average of 100 permutations of the
data was used as a baseline. b) The p-value of each eigenvalue was plotted and
the max eigenvalue with a pvalue < 0.05 was selected. c) The resulting MCA was
projected down to 3 dimensions for the purpose of density based clustering (only
the first 2 dimensions are shown). Each dot represents a tumor d) HDBSCAN was
used to perform hierarchical density based clustering tuned to minimize the
fraction of unassigned tumors. e) HDBSCAN clustered tumors were visualized on
the UMAP projection; tumors are colored by cluster membership (see legend). This
results in some tumors remaining unclassified (grey points). f) kNN was used to
assign membership to the remaining tumors.

@@

```{r child='survival-methods.Rmd'}
```

>>>>

We investigated this dataset further using multiple correspondence analysis
(MCA), and visually summarized the analysis with UMAP (**figure 2**). We used
the resulting UMAP graph coordinates to perform density based clustering with
HDBSCAN, which resulted in identification of 10 well-defined clusters capturing
about 80% of the tumor samples. To capture the remaining samples into one of
these 10 clusters we used kNN (see **Supplementary Methods** for details on all
clustering algorithms).

@@

```{r, child='figure2.Rmd'}
```

@@

We used Bayesian inference to explore the relative impact of cancer diagnosis on
survival at the time of diagnosis. Our model assumes that at birth each person
has a small initial probability per unit time to die, and that this probability
grows exponentially in time at some baseline rate that matches the observed CDC
data. The rate of growth in risk is assumed to jump to a higher level at
diagnosis, which we take as a proxy for onset. This represents an accelerated
aging type model. To fit the model, we constrain the baseline risks using the
available CDC life tables. We then constrain the cancer specific multiplier
using the longevity data within the cancer tumor dataset. Our model estimates
the effect of cancer type and class-specific cancer effects independently,
resulting in a cancer and class specific estimate of the aging rate multiplier.
A detailed description of the CDC longevity model and likelihood is provided in
the supplementary methods.
